Graph Topological Aspects of Granger Causal Network Learning

### 2019-07-05 paper/paper.tex
Some updates on paper prose.

@@

A second and wildly successful heuristic is the LASSO regularizer
\cite{tibshirani1996regression}, which can be understood as a natural
convex relaxation to penalizing the count of the non-zero edges.  The
LASSO enjoys fairly strong theoretical guarantees
\cite{wainwright2009sharp}, extending largely to the case of
stationary time series data with a sufficiently fast rate of
dependence decay \cite{basu2015} \cite{wong2016lasso}
\cite{autoregressive_process_modelling_via_the_lasso_procedure}, and
variations on the LASSO have been applied in a number of different
time series contexts as well as Granger-causality
\cite{DBLP:journals/corr/HallacPBL17} \cite{haufe2008sparse}
\cite{bolstad2011causal} \cite{he2013stationary}
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery}.
One of the key improvements to the original LASSO algorithm is the
adaptive (i.e. weighted) ``adaLASSO'' \cite{adaptive_lasso_zou2006},
for which oracle results (i.e. asymptotic support recovery) are
established under less restrictive conditions than for the vanilla
LASSO \cite{wainwright2009sharp}.%   Success of the LASSO has lead to
% the famous ``bet on sparsity'' principle ``Use a procedure that does
% well in sparse problems, since no procedure does well in dense
% problems.''  \cite{tibshirani2015statistical}.

@@

The principle contributions of this paper are as follows: firstly, in
section \ref{sec:theory} we study \textit{pairwise} Granger-causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as a characterization of the sparsity pattern of matrices arising
in the Wold decomposition, generalizing in some sense the notion of
``feedback-free'' processes studied by \cite{caines1975feedback} in
close connection with Granger-causality.  We establish sufficient
conditions (sections \ref{sec:strongly_causal_graphs},
\ref{sec:persistent_systems}) under which a fully conditional
Granger-causality graph can be recovered from pairwise tests alone
(sec \ref{sec:pairwise_algorithm}).  Secondly, we propose in section
\ref{sec:structure_learning} a graph search heuristic which implements
our theoretical results to finite data samples, specifying and
summarizing appropriate methods for hypothesis testing (section
\ref{sec:pairwise_hypothesis_testing}), model order selection (section
\ref{sec:model_order_selection}), computationally efficient estimation
(section \ref{sec:efficient_model_estimation}), and error rate
controls (section \ref{sec:error_rate_control}).  Our heuristics are
compared against the LASSO algorithm in Section
\ref{sec:empirical_evaluation}.  We stress the scalability of our
algorithm which is capable of comfortably handling hundreds or
thousands of nodes on a single machine, as opposed to standard LASSO
algorithms which do not take advantage of the special structure
associated with stationary time series data.  In section
\ref{sec:application} we develop an example application.  And provide
concluding remarks on further open problems in our concluding section
\ref{sec:conclusion}.  \todo{do an application}

@@

\section{Theory}
\label{sec:theory}
\subsection{Formal Setting}
Consider the space $L_2(\Omega)$, the usual Hilbert space of finite
variance random variables over a probability space
$(\Omega, \mathcal{F}, \mathbb{P})$ having inner product
$\inner{x}{y} = \E[xy]$.  We will work with a discrete time and
wide-sense stationary (WSS) vector valued process $x(t)$ (with
$t \in \Z$) with elements taking values in $L_2$.  We suppose that
$x(t)$ has zero mean, $\E x(t) = 0$, and has absolutely summable
matrix valued covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$ with an absolutely
continuous spectral density $S(\omega)$:

@@

\begin{equation}
  \label{eq:hilbert_space_defn}
  \begin{aligned}
    % \H^x &= \cl \{\sum_{\tau = -T}^T a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, T \in \N\} \subseteq L_2(\Omega),\\
    \H_t^x &= \cl \{\sum_{\tau = 0}^T a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, T \in \N\} \subseteq L_2(\Omega)\\
    H_t^x &= \{a x(t)\ |\ a \in \R\} \subseteq L_2(\Omega),
  \end{aligned}
\end{equation}

where the closure is naturally in mean-square.  We will often omit the
superscript $x$ which should be clear from context.  Evidently these
spaces are separable, and as closed subspaces of a Hilbert space they
are themselves Hilbert.  We will denote the spaces generated in
analogous ways by particular components of $x$ as e.g.
$\H_t^{(i, j)}$, $\H_t^{i}$ or by all but particular components as
e.g. $\H_t^{-j}$.

@@

where $c(t)$ is a purely deterministic sequence\footnote{the purely
  deterministic sequence $c(t)$ is one which lies in the remote past
  $\bigcap_{\tau=1}^\infty \H_{t - \tau}^x$ of the process.  For such
  processes a single sample $c(t_0)$ is enough to determine $c(t)$ for
  every $t$.  For example,
  $c(t) = \text{sin}(2\pi t + \Theta);\; \Theta \sim \mathcal{U}[-\pi,
  \pi]$}, $v(t)$ is an uncorrelated sequence and $A(0) = I$.  We
will assume that $c(t) = 0$, which in practice is to say that it has
been accurately estimated and removed.  We additionally require that
this representation can be inverted to yield the $\VAR(\infty)$ form

@@

\begin{definition}[Basic Setup]
  The process $x(t)$ is an $n$ dimensional wide sense stationary
  process having invertible $\VAR(\infty)$ representation
  \eqref{eqn:ar_representation} where $v(t)$ is sequentially
  uncorrelated and has a diagonal covariance matrix.  The $MA(\infty)$
  representation of equation \eqref{eqn:wold} has $c(t) = 0$ and
  $A(0) = I$.
\end{definition}

@@

\begin{definition}[Granger Causality]
  \label{def:granger_causality}
  For the WSS series $x(t)$ we say that component $x_j$
  \textit{Granger-Causes} (GC) component $x_i$ (with respect to $x$)
  and write $x_j \gc x_i$ if given Hilbert spaces
  $\H_{t - 1}$, $\H^{-j}_{t - 1}$

@@

  \begin{enumerate}
    \item{$x_j \ngc x_i$}
    \item{$\forall \tau \in \N_+\ B_{ij}(\tau) = 0$ i.e. $\B_{ij}(z) = 0$}
    \item{$\mathcal{F}_{j \rightarrow i} \defeq \ln\frac{\xi_i}{\xi_{ij}} = 0$}
    \item{$\H_t^{i} \perp \H_{t - 1}^{j}\ |\ \H_{t - 1}^{-j} \iff H_t^{i} \perp \H_{t - 1}^{j}\ |\ \H_{t - 1}^{-j}$}
    \item{$\linE{x_i(t)}{\H_{t - 1}^{-j}} = \linE{x_i(t)}{\H_{t - 1}}$}
    % \item{\hl{Wold $A(\tau)$ condition}}
    % \item{Tie in the 0s of $S(\omega)^{-1}$, though this isn't immediately directed.}
  \end{enumerate}
\end{theorem}

@@

  A (directed) \textit{path} (of length $r$) between nodes $i, j$,
  denoted $\gcgpath{i}{j}$, is a sequence
  $a_0, a_1, \ldots, a_{r - 1}, a_r$ with $a_0 = i$ and $a_r = j$ such
  that $\forall\ 0 \le k \le r\ (a_k, a_{k + 1}) \in \gcge$.

  A \textit{cycle} is a path of length $2$ or more between a node and
  itself.  An edge between a node and itself $(i, i)$ (which is not a
  cycle) is referred to as a \textit{loop}.

@@

\begin{definition}[Parents, Grandparents, Ancestors]
  A node $j$ is a \textit{parent} of node $i$ if $(j, i) \in \gcge$.
  The set of all $i$'s parents will be denoted $\pa{i}$.

@@

  Note that since loops are not considered paths, it is not possible
  for a node to be it's own parent.  However, unless $\gcg$ is a DAG,
  a node can be it's own ancestor.  We will ocassionally need to
  explicitly exclude $i$ from $\anc{i}$, in which case we will write
  $\anc{i}\setminus \{i\}$.
\end{definition}

@@

An elementary theorem connecting the adjacency matrix with paths will
allow us to deduce the sparsity pattern of $\A(z)$.  We omit proof,
though it follows easily by induction:

@@

From the $\VAR$ representation of $x(t)$ there is clearly a tight
relationship between each node and it's parent nodes, the following
proposition is analogous to the definition of feedback free processes
of \cite{caines1975feedback}:

@@

This statement is ultimately about the sparsity pattern in the Wold
decomposition matrices $A(\tau)$ since
$x_i(t) = \sum_{\tau = 0}^\infty \sum_{j = 1}^n A_{ij}(\tau)v_j(t -
\tau)$.  The proposition states that if $j \not \in \anc{i}$ then
$\A_{ij}(z) = 0$, however the implication
$j \in \anc{i} \implies \A_{ij}(z)$ need not hold.

@@

\subsection{Pairwise Granger Causality}
\label{sec:pwgc}
Recall that Granger-causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  For
example, $x_k$ may be a common ancestor which when observed completely
explains the connection from $x_j$ to $x_i$.  In this paper we study
\textit{pairwise} Granger-causality, and seek to understand when
knowledge of pairwise relations is sufficient to deduce the true fully
conditional relations of $\gcg$.

@@

This notion is of interest for a variety of reasons.  From a purely
conceptual standpoint, we will see how the notion can in some sense
capture the idea of ``flow of information'' in the underlying graph.
It may also be useful for reasoning about what the conditions under
which \textit{unobserved} components of $x(t)$ may or may not
interfere with inference in the actually observed components.
Finally, motivated from a practical standpoint to analyze causation in
large systems, we will seek to construct practical estimation
procedures based purely on pairwise causality tests since the
computation of such pairwise relations is somewhat easier.

@@

\begin{proposition}
  \label{prop:ancestor_properties}
  If in a Granger-causality graph $\gcg$ we have $j \pwgc i$ then
  $j \in \anc{i}$ or $\exists k \in \anc{i} \cap\anc{j}$ which is a
  confounder of $(i, j)$.
\end{proposition}

@@

\begin{corollary}
  If the graph $\gcg$ is acyclic and if $j \pwgc i$ and $i \pwgc j$ then $\exists k \in \anc{i} \cap \anc{j}$ confounding $(i, j)$.
\end{corollary}

@@

\begin{definition}[Strongly Causal]
  \label{def:strongly_causal}
  We will say that a Granger-causality graph $\gcg$ is
  \textit{strongly causal} if there is at most 1 directed path between
  any two nodes.  Strongly Causal Graphs will be referred to as ``SCGs''.
\end{definition}

@@

In light of proposition \ref{prop:sc_graph_common_anc}, the following
provides a partial converse to proposition \ref{prop:ancestor_properties}.

@@

\begin{remark}
  As we have seen and as is true in much of statistics, confounding
  nodes pose challenges for Granger-causality.  However, as opposed to
  Pearl's causal calculus \cite{pearl2000art}, pairwise
  Granger-causality does not suffer any difficulty with so-called
  ``colliders'', that is, the topology $i \rightarrow k \leftarrow j$
  will never result in $i \pwgc j$ or $j \pwgc i$.  This is evidently
  an advantage of the \textit{temporal} nature of Granger causality --
  ``information'' cannot flow backwards along the edges of $\gcg$.
\end{remark}

@@

  i.e. the ``first'' and ``last'' coefficients of the filter $\B(z)$,
  where $\tau_\infty(\B) \defeq \infty$ if the filter has an infinite
  impulse response.
\end{definition}

@@

\begin{remark}
  In the context of Granger-causality, ``most'' systems should be
  persistent.  In particular, $\mathsf{\VAR}(p)$ models are likely to
  be persistent since these naturally result in equivalent an
  $\mathsf{MA}(\infty)$ representation, except for the pathological
  case where $\B(z)$ is Nilpotent.

  Moreover, persistence is not the weakest condition necessary for the
  results of this section, but we opt to avoid unnecessarily
  complicated definitions.
\end{remark}

\begin{example}
  process $x(t)$ generated by the $\VAR(1)$ model\footnote{Recall that
    any $\VAR(p)$ model with $p < \infty$ can be written as a
    $\VAR(1)$ model, so we lose little generality in considering this
    case.}  having $\B(z) = Bz^{-1}$ where $B$, we will examine
  conditions that guarantee persistence.  Pick any
  $i \in [n], j \in \anc{i} \setminus \{i\}$, then the stability of
  $B$ allows us to write

@@

  \begin{equation*}
    \Dg(\lambda)^r \lambda = Vz
  \end{equation*}

  had a solution in $z$ for every $r \ge N$, where $V$ is an
  $n \times n - 1$ full-rank matrix spanning the nullspace of $\beta$.
  That is, iterates of $\Dg(\lambda)$ applied to $\lambda$ would need
  to remain inside $\mathcal{N}(\beta)$.  This would imply that

  \begin{equation*}
    VV^\dagger \lambda^{r + 1} = \lambda^{r + 1},
  \end{equation*}

  i.e. that $\lambda^{r + 1}$ is an eigenvector of $VV^\dagger$ for an
  infinite number of integers $r$.  However, since there can only be a
  finite number of (unit length) eigenvectors, this cannot be the case
  unless every eigenvalue $(\lambda_1, \ldots, \lambda_n)$ were equal.

@@

We populate each of the edges (including self loops) with random
linear filters constructed by placing $5$ transfer function poles
(i.e. $p = 5$) uniformly at random in a disc of radius $3 / 4$ (which
guarantees stability for acyclic graphs).  The resulting system is
driven by i.i.d. Gaussian random noise, each component having random
variance $\sigma_i^2 = 1/2 + r_i$ where $r_i \sim \text{exp}(1/2)$.
We set $p_{\text{max}} = 15$.  Results and representative graphs are
collected in Figure \ref{fig:simulation_results}.

@@

\begin{figure}
  \centering
  \caption{PWGC Compared Against AdaLASSO \cite{adaptive_lasso_zou2006} (SCG)}
  \label{fig:simulation_results_comparison1}
  \includegraphics[width=0.95\textwidth]{new_lasso_comparison_scg_simulation.pdf}

@@

\begin{figure}
  \caption{PWGC vs adaLASSO (DAG, $q = \frac{2}{n}$)}
  \label{fig:simulation_results_comparison2}
  \includegraphics[width=0.95\textwidth]{new_lasso_comparison_dag_simulation.pdf}
  {\scriptsize Figure \ref{fig:simulation_results_comparison2}
    provides results for systems which do not guarantee the
    assumptions of Theorem \ref{thm:scg_recovery}, though the graph
    has a similar level of sparsity.}
\end{figure}

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
    \caption{Fixed $T$, increasing $n$ (SCG)}
    \label{fig:simulation_results_scaling}
    \includegraphics[width=\linewidth]{new_increasing_n_simulation.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \caption{MCC Comparison for $T \le 100$}
    \label{fig:small_T_comparison}
    \includegraphics[width=\linewidth]{new_mcc_comparison000.pdf}
  \end{subfigure}

  {\scriptsize Figure \ref{fig:simulation_results_scaling} depicts
    support recovery performance as the number of nodes $n$ increases,
    and the edge proportion as well as the number of samples $T$ is
    held fixed.  Remarkably, performance appears to degrade slowly in
    $n$ -- it is the graph topology (SCG or non-SCG) as well as the
    level of sparsity (measured by $q$) which is the main factor for
    support recovery performance.

    Figure \ref{fig:small_T_comparison} provides a support recovery
    comparison for very small values of $T$ typical of some
    applications.}
\end{figure}

>>>>

% TODO: Are we, in this setting, immune to confounders?  We are removing
%       all bi-directional edges, so it seems likely that we are?  Can we state
%       any theorem about this?

@@

A second and wildly successful heuristic is the LASSO regularizer
\cite{tibshirani1996regression}, which can be understood as a natural
convex relaxation to penalizing the count of the non-zero edges.  The
LASSO enjoys fairly strong theoretical guarantees
\cite{wainwright2009sharp}, extending largely to the case of
stationary time series data with a sufficiently fast rate of
dependence decay \cite{basu2015} \cite{wong2016lasso}
\cite{autoregressive_process_modelling_via_the_lasso_procedure}, and
variations on the LASSO have been applied in a number of different
time series contexts as well as Granger-causality
\cite{DBLP:journals/corr/HallacPBL17} \cite{haufe2008sparse}
\cite{bolstad2011causal} \cite{he2013stationary}
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery}.
One of the key improvements to the original LASSO algorithm is the
adaptive (i.e. weighted) ``adaLASSO'' \cite{adaptive_lasso_zou2006},
for which oracle results (i.e. asymptotic support recovery) are
established under less restrictive conditions than for the vanilla
LASSO.  Our experimental comparisons in Section
\ref{sec:empirical_evaluation} are against the
adaLASSO. % Success of the LASSO has lead to
% the famous ``bet on sparsity'' principle ``Use a procedure that does
% well in sparse problems, since no procedure does well in dense
% problems.''  \cite{tibshirani2015statistical}.

@@

The principle contributions of this paper are as follows: firstly, in
section \ref{sec:theory} we study \textit{pairwise} Granger-causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as an interpretation in terms of the graph topology of the
sparsity pattern of matrices arising in the Wold decomposition,
generalizing in some sense the notion of ``feedback-free'' processes
studied by \cite{caines1975feedback} in close connection with
Granger-causality.  We establish sufficient conditions (sections
\ref{sec:strongly_causal_graphs}, \ref{sec:persistent_systems}) under
which a fully conditional Granger-causality graph can be recovered
from pairwise tests alone (sec \ref{sec:pairwise_algorithm}).
Secondly, we propose in section \ref{sec:structure_learning} a graph
search heuristic which implements our theoretical results to finite
data samples, specifying and summarizing appropriate methods for
hypothesis testing (section \ref{sec:pairwise_hypothesis_testing}),
model order selection (section \ref{sec:model_order_selection}),
computationally efficient estimation (section
\ref{sec:efficient_model_estimation}), and error rate controls
(section \ref{sec:error_rate_control}).  Our heuristics are compared
against the adaLASSO algorithm in Section
\ref{sec:empirical_evaluation}.  We stress the scalability of our
algorithm which is capable of comfortably handling hundreds or
thousands of nodes on a single machine, as opposed to standard LASSO
algorithms which do not take advantage of the special structure
associated with stationary time series data.  In section
\ref{sec:application} we develop an example application.  Concluding
remarks on further open problems and extentions are provided in Section
\ref{sec:conclusion}.  \todo{do an application}

@@

\section{Theory}
\label{sec:theory}
\subsection{Formal Setting}
Consider the space $L_2(\Omega)$, the usual Hilbert space of finite
variance random variables over a probability space
$(\Omega, \mathcal{F}, \mathbb{P})$ having inner product
$\inner{x}{y} = \E[xy]$.  We will work with a discrete time and
wide-sense stationary (WSS) $n$-dimensional vector valued process
$x(t)$ (with $t \in \Z$) where the $n$ elements take values in $L_2$.  We
suppose that $x(t)$ has zero mean, $\E x(t) = 0$, and has absolutely
summable matrix valued covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$ with an absolutely
continuous spectral density $S(\omega)$:

@@

\begin{equation}
  \label{eq:hilbert_space_defn}
  \begin{aligned}
    % \H^x &= \cl \{\sum_{\tau = -T}^T a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, T \in \N\} \subseteq L_2(\Omega),\\
    \H_t^x &= \cl \{\sum_{\tau = 0}^p a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, p \in \N\} \subseteq L_2(\Omega)\\
    H_t^x &= \{a x(t)\ |\ a \in \R\} \subseteq L_2(\Omega),
  \end{aligned}
\end{equation}

where the closure is naturally in mean-square.  We will often omit the
superscript $x$ which should be clear from context.  Evidently these
spaces are separable, and as closed subspaces of a Hilbert space they
are themselves Hilbert.  We will denote the spaces generated in
analogous ways by particular components of $x$ as e.g.
$\H_t^{(i, j)}$, $\H_t^{i}$ or by all but particular components as
$\H_t^{-j}$.

@@

where $c(t)$ is a purely deterministic sequence\footnote{the purely
  deterministic sequence $c(t)$ is one which lies in the remote past
  $\bigcap_{\tau=1}^\infty \H_{t - \tau}^x$ of the process.  For such
  processes a single sample $c(t_0)$ is enough to determine $c(t)$ for
  every $t$.  For example,
  $c(t) = \text{sin}(2\pi t + \Theta);\; \Theta \sim \mathcal{U}[-\pi,
  \pi]$}, $v(t)$ is an uncorrelated sequence and $A(0) = I$.  We will
assume that $c(t) = 0$, which in practice is to say that the process
has been detrended.  We additionally require that this representation
can be inverted to yield the $\VAR(\infty)$ form

@@

\begin{definition}[Basic Setup]
  \label{def:basic_setup}
  The process $x(t)$ is an $n$ dimensional wide sense stationary
  process having invertible $\VAR(\infty)$ representation
  \eqref{eqn:ar_representation} where $v(t)$ is sequentially
  uncorrelated and has a diagonal covariance matrix.  The $MA(\infty)$
  representation of equation \eqref{eqn:wold} has $c(t) = 0$ and
  $A(0) = I$.
\end{definition}

@@

\begin{definition}[Granger Causality]
  \label{def:granger_causality}
  For the WSS series $x(t)$ satisfying the assumptions of Definition
  \ref{def:basic_setup} we will say that component $x_j$
  \textit{Granger-Causes} (GC) component $x_i$ (with respect to $x$)
  and write $x_j \gc x_i$ if given Hilbert spaces $\H_{t - 1}$,
  $\H^{-j}_{t - 1}$

@@

  \begin{enumerate}
    \item{$x_j \ngc x_i$}
    \item{$\forall \tau \in \N_+\ B_{ij}(\tau) = 0$ i.e. $\B_{ij}(z) = 0$}
    \item{$F_{ij} \defeq \big(\frac{\xi_i}{\xi_{ij}} - 1\big) = 0$}
    \item{$\H_t^{i} \perp \H_{t - 1}^{j}\ |\ \H_{t - 1}^{-j} \iff H_t^{i} \perp \H_{t - 1}^{j}\ |\ \H_{t - 1}^{-j}$}
    \item{$\linE{x_i(t)}{\H_{t - 1}^{-j}} = \linE{x_i(t)}{\H_{t - 1}}$}
    % \item{\hl{Wold $A(\tau)$ condition}}
    % \item{Tie in the 0s of $S(\omega)^{-1}$, though this isn't immediately directed.}
  \end{enumerate}
\end{theorem}

@@

  A (directed) \textit{path} (of length $r$) from node $i$ to node
  $j$, denoted $\gcgpath{i}{j}$, is a sequence
  $a_0, a_1, \ldots, a_{r - 1}, a_r$ with $a_0 = i$ and $a_r = j$ such
  that $\forall\ 0 \le k \le r\ (a_k, a_{k + 1}) \in \gcge$, and where
  $(a_k, a_{k - 1})$ are \textit{distinct} for $0 \le k < r$.

  A \textit{cycle} is a path of length $2$ or more between a node and
  itself.  An edge between a node and itself $(i, i) \in \gcge$ (which
  is not a cycle) is referred to as a \textit{loop}.

@@

\begin{definition}[Parents, Grandparents, Ancestors]
  A node $j$ is a \textit{parent} of node $i$ if $(j, i) \in \gcge$.
  The set of all $i$'s parents will be denoted $\pa{i}$, and we
  explicitly exclude loops as a special case, that is,
  $i \not\in \pa{i}$ even if $(i, i) \in \gcge$.

@@

  Recall that we do not allow a node to be it's own parent, however
  unless $\gcg$ is a DAG, a node can be it's own ancestor.  We will
  ocassionally need to explicitly exclude $i$ from $\anc{i}$, in which
  case we will write $\anc{i}\setminus \{i\}$.
\end{definition}

@@

An elementary theorem connecting the adjacency matrix with paths will
allow us to deduce the sparsity pattern of $\A(z)$.  Proof follows
easily by induction:

@@

From the $\VAR$ representation of $x(t)$ there is clearly a tight
relationship between each node and it's parent nodes, the relationship
is quantified through the sparsity pattern of $B(z)$.  Similarly, the
following proposition is analogous to the definition of feedback free
processes of \cite{caines1975feedback} and provides an interpretation
of the sparsity pattern of $A(z)$ (from the MA representation of
$x(t)$) in terms of the causality graph $\gcg$.

@@

This statement is ultimately about the sparsity pattern in the Wold
decomposition matrices $A(\tau)$ since
$x_i(t) = \sum_{\tau = 0}^\infty \sum_{j = 1}^n A_{ij}(\tau)v_j(t -
\tau)$.  The proposition states that if $j \not \in \anc{i}$ then
$\A_{ij}(z) = 0$.

\todo{Do we have $A_{ij}(z) \ne 0 \implies j \in \anc{i}$?}

% TODO: ???
% , however the implication $j \in \anc{i} \implies \A_{ij}(z) \ne 0$ need not hold.

@@

\subsection{Pairwise Granger Causality}
\label{sec:pwgc}
Recall that Granger-causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  For
example, $x_k$ may be a common ancestor which when observed, completely
explains the connection from $x_j$ to $x_i$.  In this section we study
\textit{pairwise} Granger-causality, and seek to understand when
knowledge of pairwise relations is sufficient to deduce the true fully
conditional relations of $\gcg$.

@@

This notion is of interest for a variety of reasons.  From a purely
conceptual standpoint, we will see how the notion can in some sense
capture the idea of ``flow of information'' in the underlying graph.
It may also be useful for reasoning about what the conditions under
which \textit{unobserved} components of $x(t)$ may or may not
interfere with inference in the actually observed components.
Finally, motivated from a practical standpoint to analyze causation in
large systems, we will seek to construct practical estimation
procedures based purely on pairwise causality tests since the
computation of such pairwise relations is somewhat easier.
\todo{Reconsider this paragraph}

@@

\begin{proposition}
  \label{prop:ancestor_properties}
  If in a Granger-causality graph $\gcg$ where $j \pwgc i$ then
  $j \in \anc{i}$ or $\exists k \in \anc{i} \cap\anc{j}$ which is a
  confounder of $(i, j)$.
\end{proposition}

\begin{remark}
  The interpretation of this proposition is that for $j \pwgc i$ then
  there must either be ``causal flow'' from $j$ to $i$
  ($j \in \anc{i}$) or there must be a confounder $k$ through which
  common information is received.
\end{remark}

@@

\begin{corollary}
  If the graph $\gcg$ is a DAG then $j \pwgc i, i \pwgc j \implies \exists k \in \anc{i} \cap \anc{j}$ confounding $(i, j)$.
\end{corollary}

@@

\begin{definition}[Strongly Causal]
  \label{def:strongly_causal}
  We will say that a Granger-causality graph $\gcg$ is
  \textit{strongly causal} if there is at most 1 directed path between
  any two nodes.  Strongly Causal Graphs will be referred to as SCGs.
\end{definition}

@@

In light of proposition \ref{prop:sc_graph_common_anc}, the following
provides a partial converse to proposition
\ref{prop:ancestor_properties}, and supports the intuition of ``causal
flow'' through paths in $\gcg$.

@@

\begin{remark}
  As we have seen and as is true in much of statistics, confounding
  nodes pose challenges for Granger-causality.  However, as opposed to
  Pearl's causal calculus \cite{pearl2000art}, pairwise
  Granger-causality does not suffer any difficulty with so-called
  ``colliders'', that is, the topology $i \rightarrow k \leftarrow j$
  will never result in $i \pwgc j$ or $j \pwgc i$.  This is evidently
  an advantage of the \textit{temporal} nature of Granger -- there is
  no backwards causal flow along the edges of $\gcg$.
\end{remark}

@@

  i.e. the ``first'' and ``last'' coefficients of the filter $\B(z)$,
  where $\tau_\infty(\B) \defeq \infty$ if the filter has an infinite
  length.
\end{definition}

@@

\begin{remark}
  In the context of Granger-causality, ``most'' systems should be
  persistent.  In particular, $\mathsf{\VAR}(p)$ models are likely to
  be persistent since these naturally result in an equivalent
  $\mathsf{MA}(\infty)$ representation, except for the pathological
  case where $\B(z)$ is Nilpotent.

  Moreover, persistence is not the weakest condition necessary for the
  results of this section, the condition
  $\tau_0(\A_{jk}) < \tau_\infty(\A_{ik})$ for each $i, j, k$ such
  that $k \in \anc{i} \cap \anc{j}$ is enough.
  \todo{Write something on the intuition here.}
\end{remark}

\begin{example}
  process $x(t)$ generated by the $\VAR(1)$ model\footnote{Recall that
    any $\VAR(p)$ model with $p < \infty$ can be written as a
    $\VAR(1)$ model, so we lose little generality in considering this
    case.}  having $\B(z) = Bz^{-1}$, we will examine
  conditions that guarantee persistence.  Pick any
  $i \in [n], j \in \anc{i} \setminus \{i\}$, then the stability of
  $B$ allows us to write

@@

  \begin{equation*}
    \Dg(\bm{\lambda})^r \bm{\lambda} = Vz
  \end{equation*}

  had a solution in $z$ for every $r \ge N$, where $V$ is an
  $n \times n - 1$ full-rank matrix spanning the nullspace of $\beta$,
  and $\bm{\lambda} = (\lambda_1, \ldots, \lambda_n)$. That is,
  iterates of $\Dg(\bm{\lambda})$ applied to $\bm{\lambda}$ would need to remain
  inside $\beta$'s nullspace.  This would imply that

  \begin{equation*}
    VV^\dagger \bm{\lambda}^{r + 1} = \bm{\lambda}^{r + 1},
  \end{equation*}

  i.e. that $\bm{\lambda}^{r + 1}$ is an eigenvector of $VV^\dagger$
  for an infinite number of integers $r$ (the exponentiation is to be
  understood as a pointwise operation).  However, since there can only
  be a finite number of (unit length) eigenvectors, this cannot be the
  case unless every eigenvalue $(\lambda_1, \ldots, \lambda_n)$ were
  equal.

@@

We populate each of the edges (including self loops) with random
linear filters constructed by placing $5$ transfer function poles
(i.e. $p = 5$) uniformly at random in a disc of radius $3 / 4$ (which
guarantees stability for acyclic graphs).  The resulting system is
driven by i.i.d. Gaussian random noise, each component having random
variance $\sigma_i^2 = 1/2 + r_i$ where $r_i \sim \text{exp}(1/2)$.
We set $p_{\text{max}} = 15$.  Results and representative graphs are
collected in Figures \ref{fig:random_graph_topologies},
\ref{fig:simulation_results_comparison1},
\ref{fig:simulation_results_comparison2},
\ref{fig:simulation_results_scaling_and_small_T},
\ref{fig:simulation_results_dense}.

@@

\begin{figure}
  \centering
  \caption{PWGC Compared Against AdaLASSO \cite{adaptive_lasso_zou2006} (SCG)}
  \label{fig:simulation_results_comparison1}
  \includegraphics[width=0.95\textwidth]{new_lasso_comparison_scg_pmax15_simulation.pdf}

@@

\begin{figure}
  \caption{PWGC vs adaLASSO (DAG, $q = \frac{2}{n}$)}
  \label{fig:simulation_results_comparison2}
  \includegraphics[width=0.95\textwidth]{new_lasso_comparison_dag_pmax15_simulation.pdf}
  {\scriptsize Figure \ref{fig:simulation_results_comparison2}
    provides results for systems which do not guarantee the
    assumptions of Theorem \ref{thm:scg_recovery}, though the graph
    has a similar level of sparsity.}
\end{figure}

\begin{figure}
  \centering
  \caption{PWGC Scaling and Small Sample Performance}
  \label{fig:simulation_results_scaling_and_small_T}
  \begin{subfigure}[b]{0.45\textwidth}
    \caption{Fixed $T$, increasing $n$ (SCG)}
    \label{fig:simulation_results_scaling}
    \includegraphics[width=\linewidth]{new_increasing_n_simulation.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \caption{MCC Comparison for $T \le 100$}
    \label{fig:small_T_comparison}
    \includegraphics[width=\linewidth]{new_mcc_comparison001.pdf}
  \end{subfigure}

  {\scriptsize Figure \ref{fig:simulation_results_scaling} measures
    support recovery performance as the number of nodes $n$ increases,
    and the edge proportion as well as the number of samples $T$ is
    held fixed.  Remarkably, the degradation as $n$ increases is
    limited, it is primarily the graph topology (SCG or non-SCG) as
    well as the level of sparsity (measured by $q$) which are the
    determining factors for support recovery performance.

    Figure \ref{fig:small_T_comparison} provides a support recovery
    comparison for very small values of, $T$ typical for many
    applications.}
\end{figure}
